---
title: recency-surveillance
authors:
  - name: Daniel
    affiliation: The University
    roles: writing
    corresponding: true
bibliography: references.bib
---

## Section

This is a simple placeholder for the manuscript's main document [@knuth84].

```{r}
#| echo: false
#| message: false
#| results: hide
#| warning: false


library(tidyverse)
library(data.table)
library(gt)
library(sjlabelled)
library(sjmisc)
library(lubridate)
library(epiR)
library(epitools)
library(gtsummary)
library(freqtables)
```

```{r}
#| echo: false
#| message: false
#| results: hide
#| warning: false

data <- read.csv("data/fact_recency_case.csv") |>
  mutate(enrollement_date = lubridate::dmy(date_of_enrollment),
         enroll_year = lubridate::year(enrollement_date),
         enroll_month = lubridate::month(enrollement_date, label = TRUE),
         enroll_quarter = lubridate::quarter(enrollement_date)) |>
  filter(enrollement_date  >= "2021-04-01" & enrollement_date <= "2024-03-31") # up to March 31st 2024?
```

```{r}
#| echo: false
#| message: false
#| results: hide
#| warning: false

# Create and label the data frame
clean_data <- data %>%
  transmute(
    age = age_at_enrolment,
    age_group = cut(age, 
                    breaks = c(0, 14, 19, 24, 29, 120),
                    labels = c("<15", "15-19", "20-24", "25-29", "30+")
                   ),
    sex = factor(case_when(
      sex == "Female" ~ "Female",
      sex == "Male" ~ "Male",
      TRUE ~ NA_character_
    )),
    pregnant = factor(case_when(
      sex == "Female" & pregnancy_status == "No" ~ "Not pregnant",
      sex == "Female" & pregnancy_status == "Yes" ~ "Pregnant",
      sex == "Male" ~ NA_character_,
      TRUE ~ NA_character_
    )), 
    testing_history = factor(case_when(
      tested_hiv_before_encounter_date == "Yes" ~ "Ever tested",
      tested_hiv_before_encounter_date == "No" ~ "Never tested",
      TRUE ~ NA_character_
    )), 
    since_last_test = factor(case_when(
      last_tested_hiv_before_encounter == "Tested <6 months ago" ~ "<6 months",
       last_tested_hiv_before_encounter == "Tested 6-12 months ago" ~ "6-12 months",
      last_tested_hiv_before_encounter == "Tested 13-24 months ago" ~ "13-24 months",
      last_tested_hiv_before_encounter == "Tested more than 2 years ago" ~ ">2 years",
      tested_hiv_before_encounter_date ==  "No" ~ "Never tested",
      TRUE ~ NA_character_),
      levels = c("<6 months", "6-12 months", "13-24 months", ">2 years", "Never tested")),
    
    prep_use = factor(case_when(
      taking_or_ever_taken_prep == "No" ~ "Never",
      taking_or_ever_taken_prep %in% c("Yes, ever taken PrEP", "Yes, currently on PrEP") ~ "Ever/ Currently",
      TRUE ~ NA_character_
    )),
    
    modality = factor(case_when(modality == "OPD" ~ "Other PITC",
                                modality == "" ~ NA_character_,
                         TRUE ~ modality)),
    
    other_pitc = factor(case_when(modality == "Other PITC" ~ initial_hts_testing_location,
                           TRUE ~ NA_character_)),
    
    pns_identified = factor(case_when(is_identified_through_pns == 0 ~ "No",
                               is_identified_through_pns == 1 ~ "Yes",
                               TRUE ~ NA_character_)),
    
    recency_result = factor(case_when(
      final_recency_result == "Longterm" ~ "Longterm",
      final_recency_result == "Recent" ~ "Recent",
      TRUE ~ NA_character_
    )),
  ) %>%
  filter(!is.na(recency_result)) %>%
  filter(age_group %in% c("15-19", "20-24", "25-29")) %>%
  droplevels()

# check age
#clean_data |> group_by(age_group) |> frq(age)
  
```

```{r}
#| echo: false
#| message: false
#| results: hide
#| warning: false

library(sjlabelled)


# Apply labels
clean_data <- clean_data %>%
  mutate(
    age =  sjlabelled::set_label(age, "Age at Enrollment"),
    age_group = sjlabelled::set_label(age_group, "Age Group"),
    sex = sjlabelled::set_label(sex, "Gender"),
    pregnant = sjlabelled::set_label(pregnant, "Pregnancy status"),
    testing_history = sjlabelled::set_label(testing_history, "Testing History"),
    since_last_test = sjlabelled::set_label(since_last_test, "Time since last test"),
    prep_use = sjlabelled::set_label(prep_use, "PrEP use"),
    modality = sjlabelled::set_label(modality, "Testing modality"),
    other_pitc = sjlabelled::set_label(other_pitc, "Breakdown of other PITC"),
    pns_identified = sjlabelled::set_label(pns_identified, "Identified throgh PNS"),
    recency_result = sjlabelled::set_label(recency_result, "Recency result")
  ) 



```

The table reveals statistically significant differences between long-term and recent groups in terms of age distribution, time since the last test and PrEP use, while gender, pregnancy status, and testing history do not show significant differences.

```{r}
#| fig-width: 10

# Generate GT table
table_one <- clean_data %>%
  select(age_group, sex, pregnant, testing_history, since_last_test, prep_use, recency_result) %>%
  tbl_summary(
    by = recency_result,
    statistic = list(all_continuous() ~ "{median} ({sd})", all_categorical() ~ "{n} ({p}%)"),
    digits = all_continuous() ~ 2,
    missing = "no"
  ) %>%
  add_overall() %>%
  add_p() %>%
  modify_header(label ~ "**Characteristic**") %>%
  modify_spanning_header(c("stat_1", "stat_2") ~ "**Final recency result**") %>%
  as_gt() %>%
  tab_style(
    style = list(cell_borders(sides = "all", color = "gray", weight = px(1)),
                 cell_text(color = "black")),
    locations = cells_body()
  ) %>%
  tab_style(
    style = list(
      cell_text(weight = "bold", color = "black"),
      cell_fill(color = "#FFFDD0")
    ),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(
    style = list(
      cell_text(weight = "bold", color = "black"),
      cell_fill(color = "#FFFDD0")
    ),
    locations = cells_column_spanners()
  ) %>%
  opt_table_font(
    font = list(
      google_font(name = "Roboto Condensed")
    )
  ) |>
  tab_options(data_row.padding = px(0.5))

gtsave(table_one, "output/table_1.png")

table_one
```

```{r}
# Save the gt table as HTML
gtsave(table_one, "table_1.html")
gtsave(table_one, "output/table_1.html")

# Open the HTML file and copy the table
browseURL("output/table_1.html")

```

## Summarize by indicator

## Age Analysis

```{r}

rita_data <- clean_data %>%
      freq_table(age_group, recency_result, percent_ci = 95, drop = TRUE) |>
      freq_format(
        recipe = "percent_row (lcl_row - ucl_row)",
        name = "percent_95",
        digits = 2
  ) |>
  select(row_var, row_cat, col_cat, n, n_row, percent_95) |>
  mutate(across(where(is.numeric), as.character)) |>
  pivot_longer(cols = n : percent_95) |>
  pivot_wider(names_from = c("col_cat", "name"),
              values_from = value,
              names_glue = "{col_cat}_{name}") |>
  select(row_var, row_cat, Recent_n, Longterm_n, Recent_n_row, Recent_percent_95) |>
  arrange(desc(row_cat))


# Summarize
df <- clean_data %>%
      freq_table(age_group, recency_result, percent_ci = 95, drop = TRUE) |>
  select(row_cat, col_cat, n) |>
  pivot_wider(names_from = col_cat, values_from = n) |>
  rename(age_group = row_cat)

# Step 2: Convert the dataframe to a matrix (excluding the 'Name' column)
df_matrix <- as.matrix(df[, -1])

# Set row and column names
rownames(df_matrix) <- df$age_group
colnames(df_matrix) <- colnames(df)[-1]

riskratios_all <- riskratio(df_matrix, rev="rows") 
riskratios <- rownames_to_column(as.data.frame(riskratios_all$measure),
                                 var = "var") |>
  transmute(var = var,
            cpr = paste0(round((estimate),2), "(", round((lower),2), " - ", round((upper),2) ,")"))

# Bring it together
result_var <- rita_data |>
  left_join(riskratios, by = c("row_cat" = "var"))

```

```{r}
library(dplyr)
library(tidyr)
library(epitools)

process_data <- function(clean_data, variable_name) {
  # Dynamically use the variable_name in freq_table
  rita_data <- clean_data %>%
    filter(!is.na(!!sym(variable_name))) |>
    freq_table(!!sym(variable_name), recency_result, percent_ci = 95, drop = TRUE) %>%
    freq_format(
      recipe = "percent_row (lcl_row - ucl_row)",
      name = "percent_95",
      digits = 2
    ) %>%
    select(row_var, row_cat, col_cat, n, n_row, percent_95) %>%
    mutate(across(where(is.numeric), as.character)) %>%
    pivot_longer(cols = n:percent_95) %>%
    pivot_wider(
      names_from = c("col_cat", "name"),
      values_from = value,
      names_glue = "{col_cat}_{name}"
    ) %>%
    select(row_var, row_cat, Recent_n, Longterm_n, Recent_n_row, Recent_percent_95) %>%
    arrange(desc(row_cat))
  
  # Summarize data
  df <- clean_data %>%
    filter(!is.na(!!sym(variable_name))) |>
    freq_table(!!sym(variable_name), recency_result, percent_ci = 95, drop = TRUE) %>%
    select(row_cat, col_cat, n) %>%
    pivot_wider(names_from = col_cat, values_from = n) %>%
    rename(variable_group = row_cat)
  
  # Convert the dataframe to a matrix (excluding the 'variable_group' column)
  df_matrix <- as.matrix(df[, -1])
  
  # Set row and column names
  rownames(df_matrix) <- df$variable_group
  colnames(df_matrix) <- colnames(df)[-1]
  
  # Calculate risk ratios
  riskratios_all <- riskratio(df_matrix, rev="rows")
  riskratios <- rownames_to_column(as.data.frame(riskratios_all$measure), var = "var") %>%
    transmute(var = var,
              cpr = paste0(round((estimate), 2), "(", round((lower), 2), " - ", round((upper), 2), ")"),
              estimate = estimate,
              lower = lower,
              upper = upper)
  
  # Combine results
  result_var <- rita_data %>%
    left_join(riskratios, by = c("row_cat" = "var"))
  
  return(result_var)
}


```

```{r}

# Example usage:
age_group_r <- process_data(clean_data, "age_group")
sex_r <- process_data(clean_data, "sex")
pregnant_r <- process_data(clean_data, "pregnant")
testing_history_r <- process_data(clean_data, "testing_history")
since_last_test_r <- process_data(clean_data, "since_last_test")
prep_use_r <- process_data(clean_data, "prep_use")
#modality_r <- process_data(clean_data, "modality")
#other_pitc_r <- process_data(clean_data, "other_pitc")
#pns_identified_r <- process_data(clean_data, "pns_identified")
```

```{r}
combined_results <- bind_rows(
  age_group_r,
  sex_r,
  pregnant_r,
  testing_history_r,
  since_last_test_r,
  prep_use_r
) |>
  mutate(cpr = case_when(
    str_detect(cpr, "NA") ~ "Ref",  # Check for the string "NA"
    TRUE ~ cpr
  ))

```


```{r}
#| fig-width: 10

# Generate GT table
table_two <- combined_results %>%
  select(row_var : cpr) %>%
  mutate(row_var = case_match(row_var,
                              "age_group" ~ "Age",
                              "sex" ~ "Gender",
                              "pregant" ~ "Pregnancy Status",
                              "testing_history" ~ "Testing History",
                              "since_last_test" ~ "Time since last test",
                              "prep_use" ~ "PrEP use",
                              .default = row_var))|>
  gt(
    groupname_col = "row_var"
  )|>
  tab_style(
    style = list(cell_borders(sides = "all", color = "gray", weight = px(1)),
                 cell_text(color = "black")),
    locations = cells_body()
  ) %>%
  tab_style(
    style = list(
      cell_text(weight = "bold", color = "black"),
      cell_fill(color = "#FFFDD0")
    ),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(
    style = list(
      cell_text(weight = "bold", color = "black"),
      cell_fill(color = "#FFFDD0")
    ),
    locations = cells_column_spanners()
  ) %>%
  opt_table_font(
    font = list(
      google_font(name = "Roboto Condensed")
    )
  ) %>%
  cols_label(
    row_cat = "Characteristic",
    Recent_n = "# of RITA Recent",
    Longterm_n = "# of RITA Longterm",
    Recent_n_row = "Valid RTRI results",
    Recent_percent_95 = "% Recent (95% CI)",
    cpr = "Crude Prevalence Ratio (95% CI)"
  ) %>%
  tab_options(data_row.padding = px(1))

gtsave(table_two, "output/table_2.png")

table_two
```

```{r}
# Save the gt table as HTML
gtsave(table_two, "table_2.html")
gtsave(table_two, "output/table_2.html")

# Open the HTML file and copy the table
browseURL("output/table_2.html")

```





```{r}
df <- combined_results |>
  transmute(Characteristics = case_match(row_var,
                              "age_group" ~ "Age",
                              "sex" ~ "Sex",
                              "pregant" ~ "Pregnancy status",
                              "testing_history" ~ "Testing history",
                              "since_last_test" ~ "Time since last test",
                              "prep_use" ~ "PrEP Use",
                              .default = row_var),
            strata = row_cat,
            RITARecent = Recent_n,
            RITALongterm = Longterm_n,
            NewDiagnoses = Recent_n_row,
            PercentRITARecent = Recent_percent_95,
            PrevalenceRatio = cpr,
            cPR = estimate,
            lcl = lower,
            ucl = upper
            ) 

level_order <- c('Age', 'Sex', 'Pregnancy status', 'Testing history', 'Time since last test', 'PrEP Use')

prev_ratios <- df %>% 
  #filter(Group=="Total") 
  mutate(
  Characteristics = fct_relevel(Characteristics, "Age", "Sex", "Pregnancy status", "Pregnant by Age Group", "Non-pregnant by Age Group", "Testing history", "Time since last test", 'PrEP Use', "Testing modality", "Identified through PNS"),
  strata = fct_reorder(strata, cPR)
) %>%
  ggplot(aes(x = factor(strata), y = cPR, group = factor(Characteristics, levels = level_order))) +
  geom_point(aes(colour = factor(Characteristics, levels = level_order)), position = position_dodge(width = 0.5), size = 3) +
  geom_errorbar(aes(colour = factor(Characteristics, levels = level_order), ymin = lcl, ymax = ucl), position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 1) +
  scale_y_continuous(expand = c(0, 0), limit = c(0,5,1)) +
  coord_flip() +
  facet_grid(rows = vars(Characteristics), scales = "free_y", switch = "y", space = "free_y") +
  theme(
    plot.margin = margin(0.5, 0.5, 0.5, 0.5, unit = "cm"),
    plot.title = element_text(size = 15, face = "bold", hjust = 0),
    strip.text.y = element_text(size = 15),
    strip.placement = "outside",
    axis.title.x = element_text(margin = margin(t = 0.5, b = 0.5, unit = "cm"), size = 15),
    axis.title.y = element_blank(),
    axis.text = element_text(size = 15),
    legend.position = "none",
    panel.grid.major.y = element_blank(),
    strip.text.y.left = element_text(angle = 0),
  ) + labs(y = "Crude Prevalence Ratio")

print(prev_ratios)

ggsave(prev_ratios, 
       filename = "prev.png",
       width = 20, height = 20, units = "cm")
```



















# Modelling

```{r}

model_data <- clean_data |>
  select(age_group, sex, pregnant, testing_history, since_last_test, prep_use,
         recency_result) |>
  mutate(pregnant = case_when(sex == "Male" ~ "N/A",
                              TRUE ~ pregnant)) |>
  mutate(recency_result = case_when(as.numeric(recency_result) == 1 ~ 0,
                                    as.numeric(recency_result) == 2 ~ 1,
                                    TRUE ~ NA_integer_)
                                    ) |>
  drop_na()

dim(model_data)

```

```{r}
model_data |> freq_table(recency_result)
```

```{r}

# Change Reference categories
model_data$age_group <- relevel(factor(model_data$age_group), ref = "25-29") 
model_data$pregnant <- relevel(factor(model_data$pregnant), ref = "Not pregnant")
model_data$testing_history <- relevel(factor(model_data$testing_history), ref = "Never tested")
model_data$since_last_test <- relevel(factor(model_data$since_last_test), ref = "Never tested")

```


# Adjusted Risk Ratios
# Using Robust Poisson Regression

```{r}


set.seed(123)  # Set a fixed random seed for reproducibility

fit <- glm(recency_result ~ age_group + sex + pregnant + testing_history +
             since_last_test + prep_use,
           family = poisson(link = "log"),
           data = model_data)
```

```{r}
# Calculate robust standard errors
library(sandwich)
library(lmtest)

robust_se <- sqrt(diag(vcovHC(fit, type = "HC0")))
```

```{r}
# Extract coefficients and calculate risk ratios
coef <- summary(fit)$coefficients
risk_ratios <- exp(coef[, 1])
ci_lower <- exp(risk_ratios - 1.96 * robust_se)
ci_upper <- exp(risk_ratios + 1.96 * robust_se)
```

```{r}
# Create a data frame with risk ratios and robust standard errors
result <- data.frame(
  Estimate = coef[, 1],
  "Robust SE" = robust_se,
  "Risk Ratio" = risk_ratios,
  "Lower" = ci_lower,
  "Upper" = ci_upper
)

result
```

```{r}
summary(fit) 
```


## Log-Binomial Model
```{r}
# Load necessary library
library(broom)

# Fit the log-binomial model
log_binom_model <- glm(recency_result ~ age_group + sex + pregnant + testing_history +
             since_last_test + prep_use, 
             family = binomial(link = "log"), 
             data = model_data)


```


```{r}

# Summarize the model to view coefficients
summary(log_binom_model)

```

```{r}

# Extract risk ratios (exponentiated coefficients)
risk_ratios <- exp(coef(log_binom_model)) |> as.data.frame()

# Display risk ratios
print(risk_ratios)
```



```{r}
# Extract risk ratios (exponentiated coefficients)
risk_ratios_lg <- exp(coef(log_binom_model))

# Get confidence intervals for coefficients
conf_intervals <- confint(log_binom_model)

# Exponentiate the confidence intervals to get risk ratio confidence intervals
risk_ratio_ci <- exp(conf_intervals)

# p-values 
model_summary <- summary(log_binom_model)
p_values <- coef(model_summary)[, "Pr(>|z|)"]

# Combine th
results_log_binom <- data.frame(
  Predictor = names(coef(log_binom_model)),
  Risk_Ratio = risk_ratios_lg,
  CI_Lower = risk_ratio_ci[, 1],
  CI_Upper = risk_ratio_ci[, 2]
)

results_log_binom
```



















